### 2014-10-29 contract-language.tex
Integrated formalization with Sec 4.

% \subsection{Capturing Store Semantics}
% \label{sec:store_sem}

% An important aspect of our contract classification system is that the store
% semantics is also captured using the same contract language used to describe
% application-level consistency. In this regard, similar
% to~\cite{Burckhardt2014}, we can rigorously define a wide variety of store
% semantics including those that combine any subset of session and causality
% guarantees, and multiple consistency levels. For example, a store that offers
% strong consistency is captured by the contract:

% \vspace{-1em}
% \begin{smathpar}
% \scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
% \end{smathpar}

% \noindent Similarly, a store that offers per-object causal consistency is captured by the
% contract:

% \vspace{-1em}
% \begin{smathpar}
% \ccc = \forall a.~(\hboZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff}
% \end{smathpar}

% \noindent This ability to represent store semantics and application-level consistency in
% the same language is vital to contract classification.

% \subsection{Contract Comparison}

% Our goal is to classify contracts, and map the operation they describe to
% the \emph{weakest} store-level consistency level that  nonetheless
% satisfies the contract's constraints. To this end, we need a mechanism to
% compare the ``strength'' of a contract. Let $\cv_{op}$ be a contract for a
% particular operation $op$, and $\cv_{st}$ capture a particular store
% consistency level. We would like to determine whether $op$ can be
% \emph{safely discharged} at the store consistency level $\cv_{st}$ such that
% the resulting execution does not violate $\cv_{op}$.

% Since our contracts represent axiomatic definition of program executions, let
% $\Mod{\cv}$ be the set of all executions under which contract $\cv$ is
% satisfied. If every execution $\E \in \Mod{\cv_{st}}$ is also a member of
% $\Mod{\cv_{op}}$, then $op$ can be safely discharged under the store
% consistency level $\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq
% \Mod{\cv_{op}}$. This is the model-theoretic consequence relation, written as
% $\cv_{st} \models_m \cv_{op}$.

% Observe that our contract language (stripped of its syntactic sugar) is a
% carefully chosen subset of first-order logic that is known to be
% decidable~\cite{epr}.  Since first-order logic is sound and complete
% ~\cite{completeness}, $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st}
% \Rightarrow \cv_{op}$.  Due to the decidability of our contract language, this
% implication check is automatically discharged with the help of a theorem
% prover.

@@

\vspace{-1em}
\noindent The $\Delta$ referred in the above defintion captures
assumptions about the nature of primitive relations, such as $\Rvis$
and $\Rso$ are irreflexive, and the happens-before relation $\hbZ$ is
acyclic, preventing thin-air reads.

If the contract ($\cv_{op}$) of an operation ($op$) is \emph{weaker
than} a store contract ($\cv_{st}$), then constraints expressed by the
former are implied by guarantees provided by the later. Consequently,
$\cv_{op}$ is satisfiable under $\cv_{st}$. Observe that the contracts
$\scc$, $\ccc$ and $\ecc$ are themselves totally ordered with respect
to the $\le$ relation: $\ecc \le \ccc \le \scc$.  This concurs with
the intuition that any contract satisfiable under $\ecc$ or $\ccc$ is
satisfiable under $\scc$, and any contract that is satisfiable under
$\ecc$ is satisfiable under $\ccc$. However, we are interested in the
\emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$) required
to satisfy the contract. We define the corresponding consistency level
as the \emph{consistency class of the contract}. The classification
scheme is presented formally in Figure~\ref{sem:classify}.

@@

>>>>

\subsection{Semantics}

\begin{figure}
\begin{smathpar}
\stretcharraybig
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {\eff} \in \mathtt{Effect} \qquad
  {\cv} \in \mathtt{Contract} \qquad
  \set{\eff} \in \mathtt{Effect\; Set}
}\\
\EffSoup & \in & \mathtt{EffSoup}	  & \coloneqq & \set{\eff} \\
\visZ, \soZ, &	\in & \mathtt{Relations} & \coloneqq & 
  \set{\eff}\times\set{\eff} \\
\sameobjZ		&     &  & \\
{\E} 		& \in & \mathtt{ExecState}  & \coloneqq & \Exec \\
\end{array}
\end{smathpar}

\caption{An Axiomatic Execution}
\label{sem:contracts}
\end{figure}

In this section, we formally define what it means for an execution to
satisfy a contract.

\name contracts are constraints over axiomatic definitions of program
executions.Fig.~\ref{sem:contracts} summarizes artifacts relevant to
define an axiomatic execution. We formalize an axiomatic execution as
a tuple $\Exec$, where $\EffSoup$, called the \emph{effect soup}, is
the set of all effects generated during the program execution, and
$\visZ,\soZ, \sameobjZ \subseteq \EffSoup \times \EffSoup$ are
\emph{visibility}, \emph{session order}, and \emph{same object}
relations, respectively, witnessed over generated effects at run-time.

Note that the axiomatic definition an execution ($\E$) provides
interpretations for primitive relations (eg: $\visZ$) that occur free
in contract formulas, and also fixes the domain of quantification to
set of all effects ($\EffSoup$) observed during the program execution.
As such, $\E$ is a potential model for any first-order formula ($\cv$)
expressible in our contract language. If $\E$ is indeed a valid model
for $\cv$ (expressed using model-theoretic consequence relation as $\E
\models \cv$), we say that the execution $\E$ satisfied the contract
$\cv$:
\begin{definition}
An axiomatic execution $\E$ is said to satisfy a contract $\cv$ if and
only if $E \models \cv$.
\end{definition}

@@

\vspace{-1em}
\noindent The $\Delta$ referred in the above defintion is a
conjunction of assumptions about the nature of primitive relations,
such as $\Rvis$ and $\Rso$ are irreflexive, and the happens-before
relation $\hbZ$ is acyclic. A \emph{well-formed} axiomatic execution
($\E$) is expected to satisfy these assumptions (i.e., $\E \models
\Delta$).

If the contract ($\cv_{op}$) of an operation ($op$) is \emph{weaker
than} a store contract ($\cv_{st}$), then constraints expressed by the
former are implied by guarantees provided by the later. Due to the
completeness of first-order logic\cite{completeness}, this means that
any well-formed execution ($\E$) that satisfies $\cv_{st}$ (i.e.,
$\E\models \cv_{st}$) also satisfies $\cv_{op}$ (i.e., $\E \models
\cv_{op}$). Consequently, it is safe to execute operation $op$ under
store consistency level captured by $\cv_{st}$.

Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves
totally ordered with respect to the $\le$ relation: $\ecc \le \ccc \le
\scc$.  This concurs with the intuition that any contract satisfiable
under $\ecc$ or $\ccc$ is satisfiable under $\scc$, and any contract
that is satisfiable under $\ecc$ is satisfiable under $\ccc$. However,
we are interested in the \emph{weakest} guarantee (among $\ecc$,
$\ccc$, and $\scc$) required to satisfy the contract. We define the
corresponding consistency level as the \emph{consistency class of the
contract}. The classification scheme is presented formally in
Figure~\ref{sem:classify}.

@@

\subsection{Soundness of Contract Classification}

In this section, we present a meta-theoretic result that certifies the
soundness of classification-based contract enforcement. To help us
state the result, we present an axiomatic model of the our system
described informally in Sec.~\ref{sec:sysmod}:

\begin{smathpar}
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {op} \in \mathtt{Operation} \qquad
  \tau \in \mathtt{Consistency\; Class}
}\\
\cv(\tau) & \in & \mathtt{Store\; Contract\; of \; \tau} & \coloneqq & \scc,
  \ccc, \ecc\\
{\sigma} & \in & \mathtt{Session} & \coloneqq & \cdot \ALT (op,\tau); \sigma \\
\Sigma 	& \in & \mathtt{Session\;Soup}   & \coloneqq & 
  \langle s,{\sigma} \rangle \pll \Sigma \ALT \emptyset \\
				&			&	\mathtt{Store}		  & \coloneqq & \E;\Sigma \\
\end{array}
\end{smathpar}

We model the system as a tuple $\E;\Sigma$, where the execution $\E$
captures the data store's current state of the execution, and
$\Sigma$, the session soup, is the set of concurrent client sessions
interacting with the store. A session is modeled simply as a sequence
of replicated data type operations ($op$), tagged with the consistency
class ($\tau$) of their contracts (as determined by the contract
classification scheme). We assume rewrite relation of form:

\vspace{-1.7em}
\begin{smathpar}
  \auxred{} {\E;\langle (op,\tau);\sigma \rangle \pll \Sigma} {\eff} 
    {\E';\langle \sigma \rangle \pll \Sigma'}
\end{smathpar}

\noindent on the system state. The relation captures store's progress
in execution (from $\E$ to $\E'$) due to the application of an
operation ($op$) from one of the sessions in $\Sigma$ to a replica,
generating a new effect ($\eff$). If the resultant execution ($\E'$)
satisfies the store contract ($\cv(\tau)$) of $\tau$ (i.e., $\E
\models \cv(\tau)$), the we say that the store has \emph{enforced} the
contract $\cv(\tau)$ in the execution $\E'$.

With help of the axiomatic model, we now state the soundness of
contract enforcement as following:

\begin{theorem}[Soundness of Contract Enforcement]
\label{lem:core-preservation}
Let $\cv$ be a well-formed contract of a replicated data type operation $op$,
and let $\tau$ denote the consistency class of $\cv$ as determined by
the contract classification scheme. Forall well-formed execution
states $\E$, $\E'$ such that
$\auxred{} {\E;\langle (op,\tau);\,\sigma \rangle \pll \Sigma} {\eff}
 {\E';\langle \sigma \rangle \pll \Sigma}$
if $\E' \models [\eff/\cureff]\,\cv(\tau)$, then $\E' \models [\eff/\cureff]\,\cv$
\end{theorem}

The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ contracts in all well-formed executions, then the
same store, extended with classification scheme, can enforce all
well-formed \name contracts. The proof of the theorem has been
relegated to our tech report~\cite{techrep} owing to space
constraints \footnote{Tech report also contains operational semantics of our
implementation of the data store, which concretizes the rewrite
relation ($\xrightarrow{}$), along with the proof that it correctly
enforces all \name contracts.}.
